---
title: "Getting Started: Soft Cell-Cell Communication"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: Soft Cell-Cell Communication}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Why "soft" communication scoring?

Standard ligand-receptor tools (CellChat, LIANA, CellPhoneDB) require every
cell to be assigned one discrete cell type before communication scores are
computed. That throws away exactly the information that's most interesting
in transitional biology: cells that are genuinely *between* identities, like
the partial-EMT malignant cells described in Puram et al. 2017.

CellWalker already computes a full probability distribution over cell types
for every cell via graph diffusion. This package uses that distribution
directly as a weight, instead of collapsing it to `argmax` before handing
things off to a communication tool.

## Minimal end-to-end example (synthetic data)

```{r}
devtools::load_all("..")

set.seed(1)
n_genes <- 20
n_cells <- 100
gene_names <- c("TGFB1", "TGFBR2", "VEGFA", "KDR", "CXCL12", "CXCR4",
                 "JAG1", "NOTCH1", "TNF", "TNFRSF1A",
                 paste0("Gene", 1:10))

expr_mat <- matrix(rpois(n_genes * n_cells, lambda = 2), nrow = n_genes,
                    dimnames = list(gene_names, paste0("cell", 1:n_cells)))

# Fake CellWalker output: 3 cell types, most cells confident, a few ambiguous
types <- c("TypeA", "TypeB", "TypeC")
type_probs <- matrix(0, nrow = n_cells, ncol = 3,
                      dimnames = list(colnames(expr_mat), types))
for (i in 1:n_cells) {
  if (i %% 10 == 0) {
    type_probs[i, ] <- c(0.34, 0.33, 0.33)  # ambiguous cell
  } else {
    dominant <- sample(types, 1)
    type_probs[i, dominant] <- 0.8
    type_probs[i, setdiff(types, dominant)] <- 0.1
  }
}

entropy <- identity_entropy(type_probs)
ambiguous <- flag_ambiguous_cells(entropy)

lr_pairs <- example_lr_pairs()
soft_expr <- weighted_type_expression(expr_mat, type_probs)
hard_expr <- hard_type_expression(expr_mat, type_probs)

soft_scores <- score_lr_pairs(soft_expr, lr_pairs)
hard_scores <- score_lr_pairs(hard_expr, lr_pairs)
compare_soft_hard(soft_scores, hard_scores)
```

## Is a soft score significant?

A soft LR score can be large just because a ligand and receptor are both
highly expressed overall, not because they're specifically co-expressed by
cells confidently identified as the source/target types. `soft_lr_significance()`
tests this the same way CellPhoneDB tests hard cluster scores: shuffle which
cell owns which identity-probability vector, recompute the score under that
null many times, and see how often chance beats the real value.

```{r}
sig <- soft_lr_significance(expr_mat, type_probs, lr_pairs, n_perm = 200, seed = 1)
head(sig[order(sig$p_value), c("source", "target", "ligand", "receptor", "score", "p_value", "q_value")])
```

## Real pipeline

For the actual GSE103322 analysis, run `scripts/01_download_data.R` through
`scripts/04_compare_networks.R` in order. See the repo README for the
biological motivation and dataset details.
